Ten years of school closures and consolidations in Hamilton, Canada and the impact on multimodal accessibility

### 2025-01-19 paper/Manuscript.tex
R2 submitted.

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\raggedbottom
\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={10 years of school closures and consolidations: the impact on school accessibility},
  pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
  pdfkeywords={accessibility; spatial availability; carbon emissions;
journey-to-school; walkability; active transportation; school
consolidation policy; service provision thresholds},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[10 years of school closures and consolidations: the impact on
school accessibility]{10 years of school closures and consolidations:
the impact on school accessibility}

@@

\abstract{Reducing motorized travel and increasing active travel are top
priorities for many communities. However, these priorities clash against
many government's reluctance to invest in public services, including
schools. Public school boards are increasingly asked to do more with
less, a directive often accomplished by closing and consolidating
schools. Alas, this policy tends to reduce the proximity of schools for
many students, thus reducing the chances for active travel. In this
paper we study the City of Hamilton, a mid-sized city in Ontario,
Canada, which experienced numerous elementary schools closures between
2011 and 2021. Our analysis evaluates how school closures and
consolidations reshaped accessibility for elementary school students by
motorized and active travel. In this task we use spatial availability, a
singly-constrained multimodal measure of spatial accessibility. This
method's proportional allocation mechanisms allow us to make consistent
comparisons in accessibility between time periods. Analysis is conducted
at the parcel level, considering on-the-ground-capacity (OTGC) of
schools and observed home-to-school trip lengths for motorized and
active modes. Our findings reveal that overall spatial availability in
the city declined, and while the school consolidation policy pursued by
the province may have achieved cost savings in facility operation and
maintenance, it imposes ongoing costs on students, families, and
communities through reduced opportunities for physical activity and
entrenched reliance on motorized travel.}

@@

Public services accessible within communities tend to oscillate:
economic and demographic shifts, rural and urban configurations and
political in/action all induce change. These forces spur pressures to
reorganize and consolidate public facilities, potentially improving
access for some but not for all residents in a community (Christiaanse
2020; Rosik, Puławska-Obiedowska, and Goliszek 2021). Public schools are
a compelling case study in this respect. Across the globe, school
closure and consolidation policies have been pursued by governments to
reduce the cost of running public education systems (Rong et al. 2022;
Dai et al. 2019). Research has highlighted that these polices have
knock-on implications, including the weakening of communities' social
infrastructure (Butler, Kane, and Cooligan 2019; Irwin and Seasons 2012;
Autti and Hyry-Beihammer 2014) and negative household-level fiscal
impacts such as a reduction of property values (Merrall, Higgins, and
Páez 2024). Operationally, savings may be experienced by governments,
but they are paid at different levels by students, families,
communities, and ultimately society in general.

A school offers many things to a community, and one relevant dimension
is how it shapes travel. This dimension is often overlooked by
decision-makers tasked with implementing school closure and
consolidation policies (Bierbaum, Karner, and Barajas 2021; J. Lee and
Lubienski 2017). When schools and residential locations are co-located,
the journey-to-school presents an excellent opportunity for physical
activity (Desjardins et al. 2024). However, school closures exchange the
opportunities for active trips by students (Morency, Roorda, and Demers
2009; Morency, Demers, and Lapierre 2007) for the near certainty of
motorized travel (Rong et al. 2022). Indeed, motorized travel to school
is the largest contributor of carbon emissions from personal daily
travel after the journey-to-work (Rong et al. 2022; Pantelaki, Claudia
Caspani, and Maggi 2024). Alas, these impacts are not often considered
in policy when schools close and/or consolidate (Sageman 2022), a
relatively common occurrence globally. For instance, 65\% of
comprehensive schools (e.g.~grades 1 through 9) in rural Finland since
1990 closed (Autti and Hyry-Beihammer 2014), 10\% of elementary schools
(kindergarten through grade 8) in Chicago, USA in 2013 (J. Lee and
Lubienski 2017), and 7\% of primary and lower secondary schools
(kindergarten through grade 9) were closed in Denmark between 2010-2011
(Beuchert et al. 2018).

In this paper, we investigate the case of Hamilton, a mid-sized city in
Canada that has been forced by the province of Ontario to pursue school
closure and consolidation policies. We quantify the transportation
implications of these policies using the capacity of schools in 2011,
2016 and 2021, after several waves of school closures and
consolidations. To this end, we use a novel multimodal extension of a
singly-constrained accessibility measure known as spatial availability
(Soukhov et al. 2023; Soukhov et al. 2024). We use this measure to
estimate active (i.e., walking) and motorized (i.e., car) accessibility
while constraining the supply of the service to the capacity schools. In
this way, this work furnishes evidence to document how government
austerity hurts public goals (such as physical activity) and locks
school districts and families into motorized patterns of travel for the
foreseeable future.

To foreshadow the paper's conclusions, our findings indicate that school
seats have became less available in the city. This change happened
unevenly, with certain communities being disproportionately affected.
The school closure and consolidation policy pursued by the government
may have achieved cost savings in facility operation and maintenance.
However, it has come at a cost to families, students, and communities,
resulting in lost opportunities for healthier life-styles and an
inevitable increase in motorized travel (Hammer and Baluja 2012; Rose
2013; BUCK 2019). To increase the diffusion of this analysis, all work
is transparent and reproducible, following best practices in the spatial
sciences (Brunsdon and Comber 2021; Antonio Páez 2021). For this reason,
all associated code and data are publicly available within the lead
author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}.

@@

Schools are a service that has been studied across several
transportation dimensions. For instance, Marques, Wolf, and Feitosa
(2020) demonstrate that lower-income neighbourhoods in Portugal tend to
have lower accessibility to schools compared to more advantaged
neighbourhoods. Similar conclusions are drawn in West Virginia, US by
Talen (2001), England by Burgess et al. (2011), Baton Rouge, US by
Williams and Wang (2014), São Paulo and Curitiba, Brazil by
Moreno-Monroy, Lovelace, and Ramos (2018), Pizzol, Giannotti, and
Tomasiello (2021), and Bittencourt and Giannotti (2023), and Santiago,
Chile by Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
demonstrated that school closures are seldom politically neutral
decisions. For instance, spatial trends were explored in the context of
England by Pinch (1987), areas of declining school-aged populations in
Dreseden, Germany by Müller (2011), rural Finnish communities by Autti
and Hyry-Beihammer (2014), and Chicago, US by J. Lee and Lubienski
(2017).

The methods applied to measure school accessibility have also been
varied. For example, Marques, Wolf, and Feitosa (2020) present
indicators of school seats per student along with the shortest distance
from origin to schools within a school catchment. Similar methods are
also used in Talen (2001) and Burgess et al. (2011). Several authors
propose refinements to these methods; Müller (2011), for instance,
considers travel time by public transit from origins to schools, whereas
Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
different modes and multiple schools from a single origin along with
other indicators related to school quality. By contrast, Williams and
Wang (2014), J. Lee and Lubienski (2017), Moreno-Monroy, Lovelace, and
Ramos (2018), Tiznado-Aitken, Munoz, and Hurtubia (2021), and
Bittencourt and Giannotti (2023) use more complex approaches such as the
two-step floating catchment area (2SFCA). The 2SFCA is popular in
measure in evaluating access to healthcare services (Antonio Páez,
Higgins, and Vivona 2019) and yields a supply-to-demand ratio that
accounts for the population demanding a service and the quantity that is
supplied within a certain distance range (Shen 1998; Luo and Wang 2003).
The consideration of the supply of service is pertinent in the case of
schools and is acknowledged to as a limitation to using simpler
accessibility measures (Pizzol, Giannotti, and Tomasiello 2021).

In addition to competition for school-seats, theoretically, the use of
faster modes (e.g., motorized modes) offers an advantage in terms of a
wider range in school-seat choice. Though some of the works reviewed
consider multiple modes, explicit attention to the modal
\emph{advantage} has not been discussed. To this end, in this work we
use spatial availability, a singly-constrained measure of spatial
accessibility (Soukhov et al. 2023; Soukhov et al. 2024). Spatial
availability allows for the representation of the number of
`school-seats' that are spatially available to students by mode of
transportation (i.e., motorized and active travel), considering the
school-aged student population by zone. Further, spatial availability
can be represented as school-seats per capita, mathematically equivalent
to the supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et
al. (2023)), which increases the interpretability of the analysis, and
the comparability between years.

@@

Hamilton is a mid-size city (\textasciitilde570,000 pop) in the province
of Ontario, Canada, with rural, suburban and urban characteristics
(Government of Canada 2022a). The city's public English education system
is managed by the Hamilton-Wentworth District School Board (HWDSB) and
the Hamilton-Wentworth Catholic District School Board (HWCDSB). These
two boards, like many in the province have faced fiscal pressures from
the province. For example, in 2013, the HWDSB released its Long Term
Facilities Master Plan (HWDSB 2013), which nominally meant to ensure
``equitable, affordable and sustainable learning facilities''. This plan
indicated that 80\% of its elementary schools (attended by students aged
approximately 5 to 14) would be subject to evaluations, including a
``Pupil Accommodation Review'' (HWDSB 2013). The announcement promptly
led to a public outcry, since the outcome of Pupil Accommodation Reviews
historically translated into school closures and consolidations (Craggs
2013; Seasons 2014b).

School boards in Ontario rely on funding from the province which is
distributed through a formula based on pupil enrollment (Mackenzie 2018;
Irwin and Seasons 2012). However, this funding formula has demonstrably
become insufficient to maintain schools in a state of good repair
(Auditor General of Ontario 2015). Further, existing ``top-up'' programs
to assist in operation and maintenance costs (e.g., School Facilities
Operation and Renewal Grant) were entirely phased out by 2018 (TDSB
2024). Against this backdrop, the Pupil Accommodation Review guidelines
outlined a ``community-focused'' process for assessing the value that a
school provides to students, the community, and the local economy
(Seasons 2014b; Ministry of Education 2006). However in practice,
critical residents felt the outcomes of the Pupil Accommodation Reviews
were predetermined (Thompson, Collins, and Dean 2024): older schools
were slated to be closed, and a wholesome breakdown of costs to repair
and maintain schools was not made available (Kleinhuis 2013). The
process introduced unbalanced incentives for school boards: as a
concerned resident pointed at the time, ``{[}b{]}y closing three
schools, the HWDSB would have a strong case with the Ministry of
Education to receive full funding for a new school'' (Kleinhuis 2013).
The Ministry's framework promotes the closing and consolidation of
schools rather than cost-effective solutions for keeping older schools
open (Kleinhuis 2013).

Hamilton's two school boards (HWDSB and HWCDSB), which together provide
public English education to the majority of school-aged children in
Hamilton (FAO 2023), were particularly impacted by fiscal pressures.
While Accommodation Reviews were a province-wide trend during this
period, Hamilton's school boards underwent an unprecedented number of
reviews compared to boards in other Ontario municipalities. (Seasons
2014a). In the wave of Pupil Accommodation Reviews between 2013 and
2016, 12 of the 147 elementary schools closed in Hamilton. Likely in
part due to the unpopularity of school closures, in 2017, the provincial
government introduced a moratorium, which paused any scheduled Pupil
Accommodation Reviews. This created a backlog of Accommodation Reviews
that HWDSB caught up to in subsequent years, when it closed 12 more
schools between 2017 to 2021 (HWDSB 2023). Though elementary school
locations declined by 10\% between 2011-2021, HWDSB intends to conduct
future Pupil Accommodation Reviews as initially scheduled, waiting until
the end of the moratorium before dates are assigned (HWDSB 2023).

A justification for these Pupil Accommodation Reviews are operational
savings based on ``projected'' reductions in student population and
``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
These policies led to overall reductions in school capacity per student:
school seats in the city declined 5\%--but the impacts were felt
unevenly across the city. It should be noted that although closures
dominated the Pupil Accommodation Review process, some schools were
``consolidated'', that is, a school or two were closed and another
school was expanded or opened to accommodate students from closed
schools. Since school locations decreased proportionally more than
school-seats, elementary schools in Hamilton are on average larger than
they once were.

Existing research indicates that schools have a plethora of benefits not
captured by their maintenance and operational costs (Butler, Kane, and
Cooligan 2019; Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014),
and in this paper we focus on the transportation implications. For
students who live in proximity to schools, closures or consolidations
directly eliminate the potential for active trips, and consequently
physical activity. Closing a school permanently reduces the potential
for active trips and further entrenches reliance on motorized travel.
Additionally, from the perspective of modal equity, school-aged
populations with access to motorized transport gain a speed-time
advantage, providing them with greater flexibility and choice in
accessing school-seats.

@@

\begin{itemize}
\tightlist
\item
  Did spatial availability change for home-to-school trips following
  school closures and consolidations?
\item
  What is the extent of the loss in walking minutes for these trips?
\item
  How has motorized travel increased as a result?
\item
  Which communities, particularly households with a high prevalence of
  low-income and hence more likely experiencing transport poverty,
  experienced the greatest losses due to this policy?
\end{itemize}

@@

We focus on the evaluation of the impacts of school closures and
consolidations between 2011, 2016 and 2021. The outcome of interest is
the multimodal school-seat spatial availability in Hamilton. To research
this, the following data were prepared for each year: 1) school
configurations, 2) student population, prevalence of low-income
households, and residential parcel locations, and 3) estimated trip
length by mode. These data are described next.

@@

For this study, we use the locations of elementary schools and their
capacities at each of three years: 2011, 2016, and 2021.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\raggedbottom
\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Ten years of school closures and consolidations in Hamilton, Canada and the impact on multimodal accessibility},
  pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
  pdfkeywords={accessibility; spatial availability; carbon emissions;
journey-to-school; walkability; active transportation; school
consolidation policy; service provision thresholds},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Ten years of school closures and consolidations in Hamilton,
Canada and the impact on multimodal accessibility]{Ten years of school
closures and consolidations in Hamilton, Canada and the impact on
multimodal accessibility}

@@

\abstract{Reducing motorized travel and increased active travel are top
priorities for many communities. However, these goals are challenged by
pressures that make governments reluctant to re/invest in public
services, including schools. Public school boards are increasingly asked
to do more with less, a directive often accomplished by closing and
consolidating schools. Such policies can reduce the proximity of schools
for many students, thereby limiting opportunities for active travel. In
this paper, we examine the City of Hamilton, a mid-sized city in
Ontario, Canada, which saw numerous elementary schools closures between
2011 and 2021. Our analysis evaluates how school closures and
consolidations reshaped accessibility for elementary school students,
both by motorized and active travel. We use spatial availability, a
singly-constrained multimodal measure of spatial accessibility. This
method's proportional allocation mechanisms allow us to make consistent
comparisons in accessibility between time periods. The analysis is
conducted at the parcel level, taking into account the
on-the-ground-capacity (OTGC) of schools and observed home-to-school
trip lengths for motorized and active modes. Overall, our findings
reveal a decline in school spatial availability in the city. While the
school closure and consolidation policy pursued by the province may have
achieved cost savings in facility operation and maintenance, it imposed
ongoing costs on students, families, and communities through reduced
opportunities for physical activity and entrenched reliance on motorized
travel.}

@@

Public services accessible within communities tend to oscillate:
economic and demographic shifts, rural and urban configurations and
political in/action all induce change. These forces spur pressures to
reorganize and consolidate public facilities, potentially improving
access for some but not for all residents in a community (Christiaanse
2020; Rosik, Puławska-Obiedowska, and Goliszek 2021). Public schools are
a compelling case study in this respect. Across the globe, school
closure and consolidation policies have been pursued by governments to
reduce the per-student cost of operating public education systems (Rong
et al. 2022; Dai et al. 2019). Research has highlighted that these
polices have knock-on implications, including the weakening of
communities' social infrastructure (Butler, Kane, and Cooligan 2019;
Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014) and negative
household-level fiscal impacts such as a reduction of property values
(Merrall, Higgins, and Páez 2024). Operationally, savings may be
experienced by government, but they are paid at different levels by
students, families, communities and society.

A school offers many things to a community, and one relevant dimension
is how it shapes travel. This dimension is often overlooked by
decision-makers tasked with implementing school closure and
consolidation policies (Bierbaum, Karner, and Barajas 2021; J. Lee and
Lubienski 2017). When schools and residential locations are co-located,
the journey-to-school presents an excellent opportunity for physical
activity (Desjardins et al. 2024). However, school closures exchange
these opportunities for active transportation with the near-certainty of
motorized travel (Morency, Roorda, and Demers 2009; Morency, Demers, and
Lapierre 2007; Rong et al. 2022). Indeed, motorized travel to school is
the second-largest contributor to carbon emissions from personal daily
travel, following the journey-to-work (Rong et al. 2022; Pantelaki,
Claudia Caspani, and Maggi 2024). Unfortunately, these environmental
impacts are rarely considered when school closure/consolidation policies
are implemented (Sageman 2022), and these policies are a relatively
common occurrence globally. For instance, 65\% of comprehensive schools
(e.g.~grades 1 through 9) in rural Finland since 1990 closed (Autti and
Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
primary and lower secondary schools (kindergarten through grade 9) were
closed in Denmark between 2010-2011 (Beuchert et al. 2018).

In this context, we investigate the case of Hamilton, a mid-sized city
in Canada that pursed school closure and consolidation policies
following changes to educational funding provided by the province of
Ontario. We quantify the transportation implications of these policies
using the capacity of schools in 2011, 2016 and 2021, after several
waves of school closures and consolidations. To this end, we use a novel
multimodal extension of a singly-constrained accessibility measure known
as spatial availability (Soukhov et al. 2023; Soukhov et al. 2024). We
use this measure to estimate active (i.e., walking) and motorized (i.e.,
car) accessibility while constraining the supply of the service to the
capacity schools. In this way, this work furnishes evidence to document
how government austerity hurts public goals (such as physical activity)
and locks school boards and families into motorized patterns of travel
for the foreseeable future.

To foreshadow the paper's conclusions, our findings indicate that school
seats have became less spatially available in the city. This change
happened unevenly, with certain communities being disproportionately
affected. The school closure and consolidation policy pursued by the
school board may have achieved cost savings in facility operation and
maintenance. However, it has come at a cost to families, students, and
communities, resulting in lost opportunities for healthier life-styles
and an inevitable increase in motorized travel (Hammer and Baluja 2012;
Rose 2013; BUCK 2019). To increase the diffusion of this analysis, all
work is transparent and reproducible, following best practices in the
spatial sciences (Brunsdon and Comber 2021; Antonio Páez 2021). For this
reason, all associated code and data are publicly available within the
lead author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}.

@@

Schools are a service that has been studied across several
transportation dimensions. For instance, Marques, Wolf, and Feitosa
(2020) demonstrates that lower-income neighbourhoods in Portugal tend to
have lower accessibility to schools compared to more advantaged
neighbourhoods. Similar conclusions are drawn in West Virginia, US by
Talen (2001), England by Burgess et al. (2011), Baton Rouge, US by
Williams and Wang (2014), São Paulo and Curitiba, Brazil by
Moreno-Monroy, Lovelace, and Ramos (2018), Pizzol, Giannotti, and
Tomasiello (2021), and Bittencourt and Giannotti (2023), and Santiago de
Chile by Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
demonstrated that school closures are seldom politically neutral
decisions. For instance, spatial trends were explored in the context of
England by Pinch (1987), areas of declining school-aged populations in
Dreseden, Germany by Müller (2011), rural Finnish communities by Autti
and Hyry-Beihammer (2014), and Chicago, US by J. Lee and Lubienski
(2017).

The methods applied to measure school accessibility have also been
varied. For example, Marques, Wolf, and Feitosa (2020) presents
indicators of school seats per student along with the shortest distance
from origin to schools within a school catchment. Similar methods are
also used in Talen (2001) and Burgess et al. (2011). Several authors
propose refinements to these methods; Müller (2011), for instance,
considers travel time by public transit from origins to schools, whereas
Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
different modes and multiple schools from a single origin along with
other indicators related to school quality. By contrast, Williams and
Wang (2014), J. Lee and Lubienski (2017), Moreno-Monroy, Lovelace, and
Ramos (2018), Tiznado-Aitken, Munoz, and Hurtubia (2021), and
Bittencourt and Giannotti (2023) use more complex approaches such as the
two-step floating catchment area (2SFCA). The 2SFCA is popular in
measure in evaluating access to healthcare services (Antonio Páez,
Higgins, and Vivona 2019) and yields a supply-to-demand ratio that
accounts for the population demanding a service and the quantity that is
supplied within a certain distance range (Shen 1998; Luo and Wang 2003).
The consideration of the supply of service is pertinent in the case of
schools and is acknowledged to as a limitation to using simpler
accessibility measures (Pizzol, Giannotti, and Tomasiello 2021).

In addition to competition for school-seats, theoretically, the use of
faster modes (e.g., motorized modes) offers an advantage in terms of a
wider range in school-seat choice. Though some of the works reviewed
consider multiple modes, explicit attention to the modal
\emph{advantage} has not been discussed. To this end, in this work we
use spatial availability, a singly-constrained measure of spatial
accessibility (Soukhov et al. 2023; Soukhov et al. 2024). Spatial
availability allows for the representation of the number of
`school-seats' that are spatially available to students by mode of
transportation (i.e., motorized and active travel), considering the
school-aged student population by zone. Further, spatial availability
can be represented as school-seats per capita, mathematically equivalent
to the supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et
al. (2023)). Adopting this interpretation enhances the interpretability
of the spatial availability analysis and its comparability between
years.

@@

Hamilton is a mid-size city (\textasciitilde570,000 pop) in the province
of Ontario, Canada, with rural, suburban and urban characteristics
(Government of Canada 2022a). The city's public English education system
is managed by the Hamilton-Wentworth District School Board (HWDSB) and
the Hamilton-Wentworth Catholic District School Board (HWCDSB).
Together, HWDSB and HWDCSB provide public English education to the
majority of school-aged children in Hamilton (FAO 2023). And like many
others school boards in Ontario, HWDSB and HWCDSB faced fiscal pressures
from the province. For example, in 2013, the HWDSB released its Long
Term Facilities Master Plan (LTFMP) (HWDSB 2013), which nominally meant
to ensure ``equitable, affordable and sustainable learning facilities''.
This plan indicated that 80\% of its elementary schools (attended by
students aged approximately 5 to 14) would be subject to a ``Pupil
Accommodation Review'', an unprecedented amount of Reviews relative to
other school boards (Seasons 2014a).

The announcement of these Pupil Accommodation Reviews promptly led to
public outcry, since the outcome of Pupil Accommodation Reviews
historically translated into school closures and consolidations (Craggs
2013; Seasons 2014b). The large number of scheduled Reviews was likely
driven by the existing funding pressures faced by the school boards. The
school boards rely almost entirely on provincial funding which is
allocated through a pupil-enrolment based formula (Mackenzie 2018; Irwin
and Seasons 2012). However, this funding formula has demonstrably become
insufficient to maintain schools in a state of good repair (Auditor
General of Ontario 2015). Additionally, school boards are under intense
pressure to justify their spending to the province, and the 2012 Ontario
Budget exacerbated matters by calling for school closures as a means of
achieving efficiencies (Irwin and Seasons 2012). Furthermore, existing
``top-up'' programs designed to assist with operation and maintenance
costs, such as the School Facilities Operation and Renewal Grant, were
gradually phased out and removed by 2018 (TDSB 2024). Against this
backdrop, the Pupil Accommodation Review guidelines technically outlined
a ``community-focused'' process for assessing the value that a school
provides to students, the community, and the local economy (Seasons
2014b; Ministry of Education 2006). However in practice, critical
residents felt the outcomes of the Pupil Accommodation Reviews were
predetermined (Thompson, Collins, and Dean 2024): older schools were
slated to be closed, and a wholesome breakdown of costs to repair and
maintain schools was not made available (Kleinhuis 2013). The process
introduced unbalanced incentives for school boards: as a concerned
resident summarised at the time, ``{[}b{]}y closing three schools, the
HWDSB would have a strong case with the Ministry of Education to receive
full funding for a new school'' (Kleinhuis 2013). Ontario's Ministry of
Education's framework promotes the closing and consolidation of schools
rather than cost-effective solutions for keeping older schools open
(Kleinhuis 2013).

As forewarned, Pupil Accommodation Reviews led to school closures and
consolidations. Between 2013 and 2016, 12 of the 147 elementary schools
in Hamilton were closed. Likely influenced by the unpopularity of school
closures, in 2017, the provincial government introduced a moratorium on
starting any \emph{new} Pupil Accommodation Reviews. In the subsequent
years, HWDSB worked to clear its backlog by completing the Accommodation
Reviews that had already been initiated, resulting in the closure of 12
additional schools between 2017 to 2021 (HWDSB 2023). While the number
of elementary school locations declined by 10\% between 2011-2021, HWDSB
intends to conduct future Pupil Accommodation Reviews as initially
scheduled, awaiting the end of the moratorium before scheduling new
dates (HWDSB 2023).

A justification for closures and consolidations are estimated
operational savings based on ``projected'' reductions in student
population and ``under-utilized'' school capacity (Craggs 2012; TDSB
2024; OPSBA 2024). However, city-wide, these policies led to overall
reductions in school capacity per student. While many schools where
closed and a few expanded (e.g., consolidated), the overall number of
school seats in the city declined 5\% between 2011 and 2021, and the
impacts were felt unevenly across the city.

Furthermore, existing research indicates that schools have a plethora of
benefits not captured by their maintenance and operational costs
(Butler, Kane, and Cooligan 2019; Irwin and Seasons 2012; Autti and
Hyry-Beihammer 2014), and in this paper we focus on the transportation
implications. For students who live in proximity to schools, closures or
consolidations directly eliminate the potential for active trips, and
consequently physical activity. Closing a school permanently reduces the
potential for active trips and further entrenches reliance on motorized
travel. Additionally, from the perspective of modal equity, school-aged
populations with access to motorized transport gain a speed-time
advantage, providing them with greater flexibility and choice in
accessing school-seats.

@@

\begin{itemize}
\tightlist
\item
  Did accessibility, namely spatial availability, change for
  home-to-school trips following school closures and consolidations?
\item
  Which communities, particularly those with a higher prevalence of
  low-income households, experience the greatest non-motorized (active)
  and motorized spatial availability losses?
\end{itemize}

@@

This work focuses on the evaluation of the impacts of school closures
and consolidations in Hamilton for the years 2011, 2016 and 2021. The
primary outcome of interest is the multimodal school-seat spatial
availability. To investigate this, the following data is prepared for
each studied year: 1) school configurations, 2) student population,
prevalence of low-income households, and residential parcel locations,
and 3) estimated trip length by mode of transportation. The following
three sub-sections describe this data.

@@

For this study, we use the locations of elementary schools and their
capacities for each of three studied years: 2011, 2016, and 2021.
